A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}

### 2024-01-08 tex/ESOP24/BiSig.tex
[ tex ] Minor
[ tex ] Use `\rmfamily' for the body of a definition
[ tex ] Squeeze more space
[ tex ] Rewrite the example for binding arity with colourful annotations

\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{mathpartir}

@@

\SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}
\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
\newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}

@@

%\let\theorem\relax
%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
%\let\definition\relax
%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}

\author{Liang-Ting Chen
\orcidID{0000-0002-3250-1331}
\and 
Hsiang-Shang Ko
\orcidID{0000-0002-2439-1048}}

\institute{Institute of Information Science, Academia Sinica, Taiwan}

\title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

@@

\bibliographystyle{splncs04nat}

{\bibliography{ref}

>>>>

\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{mathpartir}
\usepackage{annotate-equations}

@@

\SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}

\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}

\newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}

@@

\spnewtheorem{defn}[theorem]{Definition}{\bfseries}{\rmfamily}
\crefformat{defn}{Definition~#2#1#3}

\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\institute{Institute of Information Science, Academia Sinica, Taiwan}
\title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

@@

\bibliographystyle{splncs04nat}
\bibliography{ref}

### 2024-01-08 tex/ESOP24/S1-intro.tex
[ tex ] Rewrite the example for binding arity with colourful annotations

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.

>>>>

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is logical and computational, a decidability proof is a proof-relevant decision procedure that computes a yes-or-no answer with a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (implying the original form due to the decidability).
Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also its justification.
This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are loosely related.
